Hydras, Ordinals \& Co. \\ A library in Coq of entertaining formal mathematics

### 2021-09-24 doc/JFLA2022/paper.tex
small corrections
Proposal of a new structure.

%% Front Matter
%%
% Regular title as in the article class.
%
\title{Hydras and Cie  \\
  Variations on Kirby \& Paris' hydra battles and ordinal numbers in Coq}

@@

\institute{
Univ. Bordeaux, CNRS, Bordeaux INP, LaBRI, UMR 5800, F-33400 Talence, France \\
  \email{pierre.casteran@labri.fr}
\and
Univ. de Paris, F-75013 Paris, France
\and
to fill!
\and
MIT CSAIL, Cambridge, Massachusetts, USA
\and
Inria, Univ. de Paris, CNRS, IRIF, UMR 8243, F-75013 Paris, France
}

@@

\titlerunning{Hydras and Cie}

@@

\subsection{Vision}
The \emph{Hydra-battles} project, part of \emph{Coq-Community}~\cite{CoqCommunity} aims to be an experimental platform for the collaborative development of commented libraries of formal proofs. \emph{Coq-Community} is a community organization that we have founded in 2018 with two goals in mind: providing a solution for the long-term maintenance of interesting Coq packages, and working collaboratively on documentation projects. The \emph{Hydra-battles} project demonstrates that these two goals are not independent: interesting Coq packages can become the basis for new documentation.
%
This umbrella project now includes evolved versions of former \emph{contribs} Cantor and Additions (under the new names of Hydra-battles and Addition-chains), the Pocklington library, extracted from Russel O'Connor's Goedel \emph{contrib}\cite{OConnor05, Goedel} and a bridge to the Gaia library (\emph{contrib} of José Grimm~\cite{Gaia,grimm:hal-00911710}).
%
By following this approach of commenting interesting Coq packages, we provide new documentation content, that contributes to the diversity of the thriving Coq documentation ecosystem.

@@

\begin{theorem}
  There is no termination measure mapping hydras to a segment $[0,\mu)$ where $\mu<\epsilon_0$ for proving the termination of  all hydra battles.\label{thm3}
\end{theorem}

\TODO[Clément]{Is there a word missing above?}
Considering measures applicable to \emph{all} battles allowed us to
focus on battles where the hydra can make an arbitrary number of copies at any time, which made our proof by contradiction artificially easier.
Unfortunately, the examples  most commonly shown in the litterature
(see for instance \cite{KP82, bauer2008, BauerHydra}) 
assume that the hydra grows $n$ copies at step $n$ of the game, which is incompatible with our proof.

@@

We now prove that Theorem~\ref{thm3} still holds with these typical battles by borrowing new combinatorial results from~\cite{KS81}.
Without loss of generality, we assume the following restrictions:

@@

\begin{figure}[h]
 \noindent
 For any $\alpha$ in $[0,\epsilon_0)$ and any $i$ in $\mathbb{N}$,
   \begin{itemize}
\item If $\alpha=0$, then $H'_\alpha (i)= i$.
\item If $\alpha=\beta+1$, then 
$H'_\alpha(i)=H'_\beta(i+1)$.
\item If $\alpha$ is a limit ordinal, then 
$H'_\alpha(i) = H'_{(\canonseq{\alpha}{i+1})}(i)$.
\end{itemize}
\vspace{4pt}
\inputsnippets{HprimeDef}
\caption{The rapidly growing hierarchy $H'$ of arithmetic functions}
\label{fig:Hprime}
\end{figure}

@@

\section{Main features}

\subsection{The Coq-Community ecosystem}
\begin{todo}
Small presentation.
\end{todo}

In order to prove formally that the length of the considered
kind of battles is not given by a primitive recursive function, we used the formalization of primitive recursive functions, part
of Russell O'Connor's contribution on G\"{o}del first incompleteness theorem \cite{OConnor05, Goedel}.
For this purpose, and above all by consideration of the scientific interest of this contribution, we chose to host and maintain this work within \textit{Coq-Community}.

@@

\subsection{Documentation with Alectryon}
\begin{todo}
  Interest of Alectryon:
  \begin{itemize}
  \item Despite the frequent changes (improvements) in the Coq scripts, code inclusion and \textit{Coq}'s answers are up to date.

@@

Our proof is very different from the computational proof of Fig\ref{fig:Ex42E0}.
By definition of additive principal ordinals, 
it suffices to prove the inequality $\omega+42< \phi_0(2)$.

@@

\section{Conclusion and perspectives}

>>>>

%% Front Matter
%%
% Regular title as in the article class.
%
\title{Hydras, Ordinals \&  Co.  \\
  A library in Coq of entertaining formal mathematics}

@@

\institute{
Univ. Bordeaux, CNRS, Bordeaux INP, LaBRI, UMR 5800, F-33400 Talence, France \\
  \email{pierre.casteran@labri.fr}
\and
Univ. de Paris, F-75013 Paris, France
\and
KTH Royal Institute of Technology, Stockholm, Sweden
\and
MIT CSAIL, Cambridge, Massachusetts, USA
\and
Inria, Univ. de Paris, CNRS, IRIF, UMR 8243, F-75013 Paris, France
}

@@

\titlerunning{Hydras, Ordinals \& Co.}

@@

\subsection{Background}

\begin{itemize}
\item formalizations of mathematics in proof assistants, such as Coq, continually grow in size and scope
\begin{itemize}
\item the Mathematical Components library and the four-color theorem and the odd order theorem in Coq
\item the Mathlib library for Lean
\end{itemize}
\item to continue to be useful as examples and reusable libraries, formalizations must be well \emph{documented} and \emph{maintained} to ensure they work with new proof assistant releases \TODO[Pierre]{add evolution of styles of proof and formalization (not limited to Coq litterature)? } 
\item Lean and Mathematical Components have accompanying books, which are written mainly to onboard new library users
\item book authors faced with several design choices: literate documents, using tools to keep the book and the proof assistant synchronized
\end{itemize}

\subsection{Vision}
The \emph{Hydra-battles} project, part of \emph{Coq-Community}~\cite{CoqCommunity} aims to be an experimental platform for the collaborative development of commented libraries of formal proofs. \emph{Coq-Community} is a community organization that we have founded in 2018 with two goals in mind: providing a solution for the long-term maintenance of interesting Coq packages, and working collaboratively on documentation projects. The \emph{Hydra-battles} project demonstrates that these two goals are not independent: interesting Coq packages can become the basis for new documentation.
%
This umbrella project now includes evolved versions of former \emph{contribs} Cantor and Additions (under the new names of Hydra-battles and Addition-chains), the Pocklington library, extracted from Russel O'Connor's Goedel \emph{contrib}~\cite{OConnor05, Goedel} and a bridge to the Gaia library (\emph{contrib} of José Grimm~\cite{Gaia,grimm:hal-00911710}).
%
By following this approach of commenting interesting Coq packages, we provide new documentation content, that contributes to the diversity of the thriving Coq documentation ecosystem.

@@

\begin{theorem}
  Let $\mu$ be any ordinal strictly less than $\epsilon_0$.
  There is no function mapping hydras to the segment $[0,\mu)$, which could be used as a measure for  proving the termination of  all hydra battles.\label{thm3}
\end{theorem}

\TODO[Clément]{Is there a word missing above?}
Considering measures applicable to \emph{all} battles allowed us to
focus on battles where the hydra can make an arbitrary number of copies at any time, which made our proof by contradiction artificially easier.
Unfortunately, the examples  most commonly shown in the litterature
(see for instance~\cite{KP82, bauer2008, BauerHydra}) 
assume that the hydra grows $n$ copies at step $n$ of the game, which is incompatible with our proof.

@@

We prove now that Theorem~\ref{thm3} still holds with these typical battles by borrowing new combinatorial results from~\cite{KS81}.
Without loss of generality, we assume the following restrictions:

@@

Using $H'$s equations as rewrite rules, we can study a concrete example of hydra battle. Let us start with the hydra of figure~\ref{fig:start} and $0$ as initial step.
By a sequence of rewritings and induction proof steps, we prove that the number of steps of the considered battle is greater or equal than $2^{2^N}$, where $N=2^{70}-1$.

\begin{figure}[h]
  \centering
  \begin{tikzpicture}[very thick, scale=0.3]
  \node (h1) at (4,0){$\bullet$};
  \node (h2) at (2,2){$\bullet$};
  \node (h3) at (0,4){$\bullet$};
  \node (h4) at (1,4){$\bullet$};
  \node (h5) at (2,4){$\bullet$};
  \node (h6) at (4.5,2){$\bullet$};
  \node (h7) at (6.5,2){$\bullet$};
  \node (h8) at (8.5,2){$\bullet$};
    \draw (h1) -- (h2) ;
    \draw (h2) -- (h3) ;
    \draw (h2) -- (h4) ;
    \draw (h2) -- (h5) ;
    \draw (h1) -- (h6) ;
    \draw (h1) -- (h7) ;
    \draw (h1) -- (h8);
 \end{tikzpicture}

  \caption{The hydra associated with the ordinal $\omega^3+3$}
  \label{fig:start}
\end{figure}

\begin{figure}[h]
 \noindent
 For any $\alpha$ in $[0,\epsilon_0)$ and any $i$ in $\mathbb{N}$,
   \begin{itemize}
\item If $\alpha=0$, then $H'_\alpha (i)= i$.
\item If $\alpha$ is a limit ordinal, then 
  $H'_\alpha(i) = H'_{(\canonseq{\alpha}{i+1})}(i)$.
  \item If $\alpha=\beta+1$, then 
$H'_\alpha(i)=H'_\beta(i+1)$.
\end{itemize}
\vspace{4pt}
\inputsnippets{HprimeDef}
\caption{The rapidly growing hierarchy $H'$ of arithmetic functions}
\label{fig:Hprime}
\end{figure}

@@

\section{Integration within Coq-Community}

\subsection{Presentation}

\subsection{Interactions with other Coq-Community projects}

In order to prove formally that the length of the considered
kind of battles is not given by a primitive recursive function, we used the formalization of primitive recursive functions, part
of Russell O'Connor's contribution on G\"{o}del first incompleteness theorem~\cite{OConnor05, Goedel}.
For this purpose, and above all by consideration of the scientific interest of this contribution, we chose to host and maintain this work within \textit{Coq-Community}.

@@

% We could already talk a bit of the monorepo structure here, and
% defer some explanations on the tooling to the next section.

\section{Modernizing the infrastructure}
\subsection{Documentation with Alectryon}
\begin{todo}
  Interest of Alectryon:
  \begin{itemize}
  \item Despite the frequent changes (improvements) in the Coq scripts, code inclusion and \textit{Coq}'s answers are up to date.

@@

Our proof is very different from the computational proof of
Figure~\ref{fig:Ex42E0}.
By definition of additive principal ordinals, 
it suffices to prove the inequality $\omega+42< \phi_0(2)$.

@@

\subsection{Continuous integration and deployment, opam packaging, etc.}
\section{Conclusion and perspectives}
